---
title: "Assignment 1"
output: html_notebook
---

### Problem 1

1. Explore the overall structure of the dataset using str(). When you read the 
data, set StringAsFactors=True to let R convert String variables to factors.

```{r}
vegas = read.table("data/LasVegasTripAdvisorReviews-Dataset.csv", sep=";",
                  header=TRUE, check.names=FALSE, stringsAsFactors=TRUE)
colnames(vegas) = gsub(" ", "", colnames(vegas))
#colnames(vegas) = gsub("\\.", "", colnames(vegas))

str(vegas)

```

2. Does any of the variables have missing values? It appears no columns have
missing values.

```{r}
colSums(is.na(vegas))
apply(is.na(vegas), 2, which)
(1-mean(complete.cases(vegas)))
```


3. Draw a histogram of the score variable. Interpret the histogram:
The Score data plotted indicates a majority of reviewers ranked their stays 
above 3. A very small proportion of reviews rated their stay as a 1 or 2. 
Additionally, a histogram may not be the best way to interpret score data 
because it limits the meaning of the scores and its relation to other 
variables.

```{r}
hist(vegas$Score, breaks=c(0:5))
```


4. Find the mode of “Score” 

```{r}
t = table(vegas$Score)
names(t[t == max(t)])
```


5. Use the “quantile” function to get the quantiles for score. What is the 
median of score?

```{r}
quantile(vegas$Score, .25)
quantile(vegas$Score, .5)
quantile(vegas$Score, .75)

median(vegas$Score)
```


6. Use the ifelse function (You can get help on the syntax of this function in 
R by typing ?ifelse) to  create  a  factor  variable  “sentiment”  which  takes  
the  value  “positive”  if  score  is  4  or  5  and  “negative”  otherwise. 
You can use the method “factor” to create a factor from a character vector. 
Take a summary of “sentiment” to make sure that the frequencies of “positive” 
and “negative” categories are consistent with the frequency  of  the  values  
in  Score.



```{r}
vegas$sentiment = ifelse(vegas$Score %in% c(4,5), "positive", "negative")
summary(vegas$sentiment)

ftable(vegas$sentiment)
ftable(vegas$Score)
```


7. Use  a  statistical  test  to  determine  if  sentiment  is  associated  with any  of  the  variables:  "Pool",  "Gym","Free.Internet", "Period.of.Stay", "traverler.type"", and "hotel.stars" (Assume the significance level alpha=0.01). 
Which of these variables are associated with sentiment? 

The Chi-Square Test of Independence indicates that sentiment is associated with 
Gym, Period of Stay, and Traveler Type.

- sentiment/Pool - p-value is less than alpha, we can conclude dependence. 
Reject null.

- sentiment/Gym - p-value is greater than alpha, we cannot reject null. Can't 
conclude they are not independent.

- sentiment/Free.Internet - p-value is less than alpha, we can conclude 
dependence. Reject null.

- sentiment/Period.of.Stay - p-value is greater than alpha, we cannot reject 
null. Can't conclude they are not independent.

- sentiment/Traveler.type - p-value is greater than alpha, we cannot reject 
null. Can't conclude they are not independent.

- sentiment/Hotel.stars - p-value is less than alpha, we can conclude 
dependence. Reject null.

```{r}
library(gmodels)
# CrossTable(x = vegas$sentiment, y = vegas$Pool)
chisq.test(vegas$sentiment, vegas$Pool)
# CrossTable(x = vegas$sentiment, y = vegas$Gym)
chisq.test(vegas$sentiment, vegas$Gym)
# CrossTable(x = vegas$sentiment, y = vegas$Freeinternet)
chisq.test(vegas$sentiment, vegas$Freeinternet)
# CrossTable(x = vegas$sentiment, y = vegas$Periodofstay)
chisq.test(vegas$sentiment, vegas$Periodofstay)
# CrossTable(x = vegas$sentiment, y = vegas$Travelertype)
chisq.test(vegas$sentiment, vegas$Travelertype)
#CrossTable(x = vegas$sentiment, y = vegas$Hotelstars)
chisq.test(vegas$sentiment, vegas$Hotelstars)
```


### Problem 2

1. Explore the structure of dataset. How many numerical and categorical 
attributes are there in the dataset: 9 numeric, 1 categoric (ocean_proximitiy)

```{r}
house = read.csv("data/housing-1.csv")
str(house)
```

2. Take a summary statistics of the dataset.Does any of the variables have 
missing values? total_bedrooms has missing values

```{r}
summary(house)
apply(house, 2, function(col)sum(is.na(col)))
colMeans(is.na(house))
```

3. What percentage of the rows have missing values in one or more variables? You 
can use the function “complete.cases” to get a logical vector indicating whether 
each row is complete (has no missing value) or incomplete (has one or more 
missing values). Run ?complete.cases to get help on this function in R. Use the 
logical vector returned by complete.cases to count the number of incomplete 
cases. ~1% of the rows have missing values. 

```{r}
?complete.cases
z = apply(house, 2, function(col)complete.cases(col))
summary(z)
(1-mean(complete.cases(house)))
```

4. Remove the rows with missing values. You can use either “complete.cases” 
function or “na.omit” function for this purpose. Run ?na.omit in R to get help 
on these functions and to see some examples.

```{r}
house = house[complete.cases(house$total_bedrooms),]
```

5. Draw a histogram and boxplot of the median_house_value. What can you say 
about the shape of median_house_value? Is the median_house_value sample 
positively skewed, symmetric, or negatively skewed? median_house_value is 
right-tailed or positively skewed.

```{r}
options(scipen=999) # disable scientific notation
hist(house$median_house_value)
boxplot(house$median_house_value)
```

6. Use Q-Q plot to examine if the distribution of median_house_value is 
approximately normal? It is not. the plot should be approximately straight and
diagonal from left to right.

```{r}
options(scipen=999) # disable scientific notation
qqnorm(house$median_house_value)
```

7. Use IQR method to find outliers of the median_house_value attribute. How many outliers does this attribute have? ~1064 items are outliers.

```{r}
### House median_house_value outliers
iqr <- IQR(house$median_house_value)
q1 <- quantile(house$median_house_value, .25)
q3 <- quantile(house$median_house_value, .75)
leftvalue <- q1-1.5*iqr
rightvalue <- q3+1.5*iqr
house[house$median_house_value<leftvalue | 
           house$median_house_value>rightvalue,]
```

8.Does any of the numerical attributes have a positive/negative linear 
correlation with median_house_value? Use scatter plots and correlation matrix to 
answer this question. median_income appears to be the only positive correlation.

```{r fig1, fig.height = 8, fig.width = 10}
numeric_att = house[, c("longitude", "latitude", "housing_median_age",
                        "total_rooms", "total_bedrooms", "population",
                        "households", "median_income", "median_house_value")]
pairs(numeric_att)
cor(numeric_att)
```

9. Is there a statistically significant difference between the mean of meadian_house_value across different levels of ocean_proximity ? Use an 
appropriate plot and statistical test (with significance level alpha=0.01) to answerthis question. We can reject the null and there is a statistically 
significant difference between the means. Visually, ISLAND has a much higher 
mean value than the other levels.. INLAND has a much lower mean value than the 
other levels.

```{r fig2, fig.width=10}
levels(house$ocean_proximity)
plot(house$median_house_value~house$ocean_proximity, col="red")
oneway.test(house$median_house_value~house$ocean_proximity, data=house)
```

10. Use ifelse function to create a new factor variable “Is_Inland” which takes 
the value “Yes” if the ocean_proximity is “Inland” and No” otherwise.

```{r}
house$Is_Inland = as.factor(ifelse(house$ocean_proximity == "INLAND", 
                                   "Yes", "No"))
```

11. Use side by side box plot to visualize the relationship between “Is_Inland” 
and “median_house_value”. Is there a significant difference between the means of median_house_value for Inland vs not Inland houses? Use an appropriate 
statistical test to answer this question ( use significance level alpha=0.01). 
Since the p-value is less than the alpha we can conclude there is a statistical
difference between the means of median_house_value and Is_Inland.

```{r}
plot(house$median_house_value~house$Is_Inland, col="red")
t.test(house$median_house_value~house$Is_Inland, alternative="two.sided")
```


### Problem 3

1. Open these files and examine the data in them. Note that the files do not 
have a header and the missing values are marked by “?” character. Each file 
contains the 14 attributes described here. Load each file to a dataframe 
(remember to set na.string=”?” so that “?” is recognized as missing not a data 
value).  

```{r}
cleveland = read.table("data/processed.cleveland.data", sep=",", na.string="?",
                   col.names=c("age", "sex", "cp", "trestbps", "chol", "fbs",
                               "restecg", "thalach", "exang", "oldpeak",
                               "slope", "ca", "thal", "num"))
hungarian = read.table("data/processed.hungarian.data", sep=",", na.string="?",
                   col.names=c("age", "sex", "cp", "trestbps", "chol", "fbs",
                               "restecg", "thalach", "exang", "oldpeak",
                               "slope", "ca", "thal", "num"))
switzerland = read.table("data/processed.switzerland.data", sep=",", na.string="?",
                   col.names=c("age", "sex", "cp", "trestbps", "chol", "fbs",
                               "restecg", "thalach", "exang", "oldpeak",
                               "slope", "ca", "thal", "num"))
va = read.table("data/processed.va.data", sep=",", na.string="?",
                   col.names=c("age", "sex", "cp", "trestbps", "chol", "fbs",
                               "restecg", "thalach", "exang", "oldpeak",
                               "slope", "ca", "thal", "num"))

```

2. Use rbind   function to combine the four dataframes into one dataframe and 
manually set the column names  using colnames function.  The name of each column/attribute is described here.   

```{r}
combined = rbind(cleveland, hungarian, switzerland, va)
```

3. Explore the overall structure of the dataset. What percentage of rows have 
missing values in one or more attributes? 

```{r}
summary(combined)
apply(combined, 2, function(col)sum(is.na(col)))
colMeans(is.na(combined)) # missing values by column
mean(is.na(combined)) # total of all rows/columns
(1-mean(complete.cases(combined))) # percent rows with a missing value
```

4. Read the data description carefully. Decide which attributes are categorical 
and convert them to “factor”  using factor function (set the “labels” option 
to give meaningful names/labels to each level)

```{r}
combined$sex = factor(combined$sex, levels=c(1,0), labels=c("Male", "Female"))
combined$cp = factor(combined$cp, levels=c(1,2,3,4), 
                     labels=c("typical angina", "atypical angina", 
                              "non-anginal pain", "asymptomatic"))
combined$fbs = factor(combined$fbs, levels=c(1,0), 
                      labels=c("fbs > 120mg/dl", "fbs < 120mg/dl"))
combined$restecg = factor(combined$restecg, levels=c(0,1,2),
                          labels=c("normal", "ST-T abnormality", 
                                   "prob/def left     ventricular hypertrophy"))
combined$exang = factor(combined$exang, levels=c(1,0), 
                        labels=c("exercise induced angina", 
                                 "no exercise induced angina"))
combined$slope = factor(combined$slope, levels=c(1,2,3), 
                        labels=c("upsloping ST seg", "flat ST seg", 
                                 "downloping ST seg"))
combined$ca = factor(combined$ca, levels=c(0,1,2,3), 
                     labels("1 fluoro vessel, 2 fluoro vessel", "3 fluoro 
                            vessel", "4 fluoro vessel"))
combined$thal = factor(combined$thal, levels=c(3,6,7),
                       labels=c("normal", "fixed defect", "reversable defect"))
```

5.  What is the median and mode of the age attribute. 54 and 54

```{r}
median(combined$age)
t = table(combined$age)
names(t[t == max(t)])
```

6. Are female patients under-represented compared to male patients in this 
dataset? Yes

```{r}
table(combined$sex)
```

7. Use “ifelse” and “factor” functions to create a new factor variable (call it  
“diagnosis”) which takes the value “No” if column 14 has the value zero and 
“Yes” otherwise.  Replace column 14 of your dataframe with this new variable.

```{r}
combined$diagnosis = factor(ifelse(combined$num == 0, "No", "Yes"))
combined[,"num"] = NULL
```

8. Explore the relationship between “diagnosis” variable and all other 13 
attributes in the dataset. Which variables are associated with “diagnosis” use 
appropriate plots and statistical tests to answer this question.

p-value is significantly smaller than the critical value in all variables.
Association: age, sex, cp, fbs, restecg, exang, slope, ca, and thal, trestbps, 
chol, thalach, oldpeak
 - 

```{r}
library(gmodels)
attach(combined)
options(scipen=999)
# Numeric
plot(age~diagnosis)
plot(trestbps~diagnosis)
plot(chol~diagnosis)
plot(thalach~diagnosis)
plot(oldpeak~diagnosis)

# 2-level test on diagnosis
t.test(age~diagnosis, alternative="two.sided")
t.test(trestbps~diagnosis, alternative="two.sided")
t.test(chol~diagnosis, alternative="two.sided")
t.test(thalach~diagnosis, alternative="two.sided")
t.test(oldpeak~diagnosis, alternative="two.sided")

# Categoric
table(diagnosis, sex)
chisq.test(table(diagnosis, sex))
table(diagnosis, cp)
chisq.test(table(diagnosis, cp))
table(diagnosis, fbs)
chisq.test(table(diagnosis, fbs))
table(diagnosis, restecg)
chisq.test(table(diagnosis, restecg))
table(diagnosis, exang)
chisq.test(table(diagnosis, exang))
table(diagnosis, slope)
chisq.test(table(diagnosis, slope))
table(diagnosis, ca)
chisq.test(table(diagnosis, ca))
table(diagnosis, thal)
chisq.test(table(diagnosis, thal))


```
